\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-16 paper.tex
Various edits

\section{Locally unary nodes}
\label{sec-locally-unary-edges}
We can define the arity of a node as the number of child nodes it has.
There is an
important distinction between the number of children a node has
in the graph and the number of children it has in the local trees,
because a (graph) child node is only a child in the local tree for position
$x$ if $x$ is in the inheritance interval for the corresponding edge.
Thus, for a given node $u$ and a genome position $x$ the arity
of $u$ in that local tree is less than or equal to the graph arity of
$u$. The arity
of a node in the local trees can therefore change as we move along
the genome.

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is hence unary in this local tree.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

@@

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer eARGs that contain
specific recombination \emph{events} for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer events but rather
infer gARGs with by differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
They contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

>>>>

\section{Locally unary nodes}
\label{sec-locally-unary-edges}
% TODO needs intro, giving some bit of warm-up for the reader.
We can define the arity of a node as the number of child nodes it has.
There is an
important distinction between the number of children a node has
in the graph and the number of children it has in the local trees,
because a (graph) child node is only a child in the local tree for position
% FIXME state this formally, as both Yan and Jere misunderstood here.
$x$ if $x$ is in the inheritance interval for the corresponding edge.
Thus, for a given node $u$ and a genome position $x$ the arity
of $u$ in that local tree is less than or equal to the graph arity of
$u$. The arity
of a node in the local trees can therefore change as we move along
the genome.

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is hence unary in this local tree.
% TODO note this is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

@@

% TODO add explanation of the bottom row.
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer eARGs that contain
specific recombination \emph{events} for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer events but rather
infer gARGs with by differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
They contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.
